---
title: "P8105 Homework 2"
author: "Jesse R. Ames"
date: "10/9/2021"
output: github_document
---

```{r}
library(tidyverse)
library(readxl)
```

## Problem 1

This problem uses the Mr. Trash Wheel dataset, available as an Excel file on the course website.

Read and clean the Mr. Trash Wheel sheet:

* specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
* use reasonable variable names
* omit rows that do not include dumpster-specific data
* round the number of sports balls to the nearest integer

```{r}
mr_trash_wheel <- read_excel(
  path = "data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx",
  sheet = "Mr. Trash Wheel",
  skip = 1, #Skip image
  range = cell_cols("A:N")) %>% #Omit extra columns
    janitor::clean_names() %>% #Make variable names nicer
    filter(!is.na(date)) %>% #Filter out monthly summary rows, grand total
    mutate(sports_balls = round(sports_balls)) #Round sports balls to the nearest integer

skimr::skim(mr_trash_wheel)
```

Read and clean precipitation data for 2018 and 2019. For each, omit rows without precipitation data and add a variable for year. Next, combine precipitation datasets and convert month to a character variable (the variable `month.name` is built into R and should be useful).

```{r}
precip_2018 <- read_excel(
  path = "data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx",
  sheet = "2018 Precipitation",
  range = "A2:B14") %>%
    mutate(year = factor(2018))

precip_2019 <- read_excel(
  path = "data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx",
  sheet = "2019 Precipitation",
  range = "A2:B14") %>%
    mutate(year = factor(2019))

#Combine, name months
precip_18_19 <- full_join(precip_2018, precip_2019) %>%
  janitor::clean_names() %>%
  mutate(month = month.name[month])

head(precip_18_19)
tail(precip_18_19)
```

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in both resulting datasets, and give examples of key variables. For available data, what was the total precipitation in 2018? What was the median number of sports balls in a dumpster in 2019?

a) The Mr. Trash Wheel dataset contains data from `r nrow(mr_trash_wheel)` dumpsters collected from May 2014 to January 2021, with details including the weight (in tons) and volume (in yards$^3$) of trash collected, and counts of the plastic bottles, polystyrene containers, cigarette butts, and several other items. Additionally, the dataset includes the number of sports balls found in each dumpster, and the number of homes in Baltimore powered by incinerating the trash in each dumpster. In 2019, the median number of sports balls in a dumpster was `r mr_trash_wheel %>% filter(year == 2019) %>% pull(sports_balls) %>% median`.

b) The precipitation dataset we created here contains the monthly total precipitation (presumably, in inches, in Baltimore) for each month in 2018 and 2019. The total precipitation in 2018 across all 12 months was `r precip_18_19 %>% filter(year == 2018) %>% pull(total) %>% sum()` inches, according to this data.

## Problem 2

This problem uses the [FiveThirtyEight data](https://www.p8105.com/dataset_fivethirtyeight.html); these data were gathered to create the interactive graphic on [this page](https://fivethirtyeight.com/features/science-isnt-broken/#part1). In particular, we’ll use the data in `pols-month.csv`, `unemployment.csv`, and `snp.csv`. Our goal is to merge these into a single data frame using year and month as keys across datasets.

First, clean the data in `pols-month.csv`. Use `separate()` to break up the variable `mon` into integer variables year, month, and day; replace month number with month name; create a president variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop`; and remove the `day` variable.

```{r}
pols_month <- read_csv(
  "data/fivethirtyeight_datasets/pols-month.csv") %>%
  separate(col = mon, into = c("year","month","day")) %>% #Separate dates into year/month/day
  mutate(month = month.name[as.integer(month)]) %>% #Name months
  mutate(prez_gop = as.factor(ifelse(prez_gop, yes = "gop", no = "dem"))) %>% #President variable
  mutate(prez = prez_gop, prez_gop = NULL, prez_dem = NULL, day = NULL) #Remove extra variables
```

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

Join the datasets by merging snp into pols, and merging unemployment into the result.

Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

Note: we could have used a date variable as a key instead of creating year and month keys; doing so would help with some kinds of plotting, and be a more accurate representation of the data. Date formats are tricky, though. For more information check out the lubridate package in the tidyverse.
